An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2017-03-29 uai2017/paper.tex
tweaks to intro and conclusions

Our test is applicable when the variance (over data samples) of the log
acceptance probability is small enough (less than 1). It's not clear at first why
this quantity should be bounded, but we will show that it is ``natural'' for
well-specified models running Metropolis-Hastings sampling with optimal
proposals~\citep{OptimalScaling01} on a full dataset. When we reduce the amount
of data for the test, the variance goes up. We have to reduce variance in one
of several ways. Either:

\begin{itemize}[noitemsep]
    \item Increase the temperature $K$ of the target distribution. Log
    likelihoods scale as $1/K$, and so the variance of the likelihood ratio will
    vary as $1/K^2$. Our model is no longer well-specified (we are doing
    inference at a temperature different from that assumed during data
    generation), but as we demonstrate in Section~\ref{ssec:logistic}, higher
    temperature can be advantageous for parameter exploration.

    \item For continuous distributions, reduce the proposal step size and
    variance compared to an optimal proposal. The variance of the log acceptance
    probability scales as the square of proposal step size.

    \item Utilize Hamiltonian and Langevin Dynamics to judiciously model a system
    so as to generate distant yet high-quality proposals.
\end{itemize}

It is worth discussing at this point the typical goals of M-H sampling on large
datasets. By the Bernstein-von Mises Theorem, the posterior distribution for a
Bayesian inference task has variance that scales inversely with $N$. Simply
sampling from it is one application, but an efficient
proposal~\citep{OptimalScaling01} has similar variance to the target and will
diffuse to it extremely slowly. For applications to neural networks or models
where the posterior is multimodal~\citep{choromanska2014loss}, samplers will
likely get trapped in one of the modes. A common solution is to anneal the
sampler, running first at high temperatures to flatten the likelihood landscape.
This in turn reduces the variance of the log acceptance probability and allows
our test to be applied. These samples can cover the search space densely with
small steps rather than taking a few sparse steps towards an optimum. In this
mode, M-H tests can be used in similar fashion to Stochastic Gradient Descent.
The goal in SGD is to make gradual progress to a posterior mode with each step,
taking small steps so that the cumulative displacement has progressively lower
variance.

The contributions of this paper are as follows:

@@

    \item We compare performance of our new test and prior approaches on several
    datasets. We demonstrate several order-of-magnitude improvements in efficiency
    (measured as data consumed per test), and show that it does not suffer from
    long-tailed minibatch sizes.
\end{itemize}

@@

The most pressing direction for future work is to integrate our M-H test with
Hamiltonian Monte Carlo and Langevin Dynamics methods to enable MCMC methods to
scale to models such as deep neural networks. With these types of models, our
proposal probabilities will almost certainly have to be asymmetric, but these
typically result in an estimate plus Gaussian noise which we can absorb into our
test.

>>>>

We note that this approach is fundamentally different from prior
work. It makes no assumptions and requires no global bounds on the
posterior parameter distribution in order to work. It is exact in the
limit as batch size increases by the Central Limit Theorem. This is
not true of ~\citep{cutting_mh_2014} and~\citep{icml2014c1_bardenet14}
which use tail bounds and provide only approximate tests even with
arbitratily large batches of data. Our test is also exact under the
assumptions of \citep{cutting_mh_2014} that per-sample parameter
posteriors are normally distributed about the population mean.  Rather
than tail bounds, our approach uses moment estimates from the data to
determine how far the minibatch posteriors deviate from a normal
distribution.  These bounds carry through to the overall accuracy of
the test.

Our test is applicable when the variance (over data samples) of the
log probability ratio between proposal and current state is small
enough (less than 1). It's not clear at first why this quantity should
be bounded, but it is natural for well-specified models running
Metropolis-Hastings sampling with optimal
proposals~\citep{OptimalScaling01} on a full dataset. If the posterior
parameter distribution is a unit-variance normal distribution, then
the posterior for $N$ samples will have variance $1/N$. There is
simply not enough information in $M << N$ samples to locate and
efficiently sample from this posterior. This is not a property of any
particular proposal or test, but of the information carried by the
data. The variance condition succinctly captures the condition that
the minibatch carries enough information to generate a sample.  While
we cannot expect to generate independent samples from the posterior
using only a small subset of the data, there are three situations
where we can exploit small minibatches:

\begin{enumerate}[noitemsep]
    \item Increase the temperature $K$ of the target distribution. Log
    likelihoods scale as $1/K$, and so the variance of the likelihood ratio will
    vary as $1/K^2$. As we demonstrate in Section~\ref{ssec:logistic}, higher
    temperature can be advantageous for parameter exploration.

    \item For continuous distributions, reduce the proposal step size
      (i.e. generate correlated samples). The variance of the log
      acceptance probability scales as the square of proposal step
      size.

  \item Utilize Hamiltonian Dynamics for proposals and tests. Here the
    dynamics itself provide shaping to the posterior distribution, and
    the M-H is only needed to correct for quantization error. In terms
    of the information carried by the samples, this approach is not
    limited by the data in a particular minibatch since momentum is
    carried over time and ``remembered'' across multiple
    minibatches.
\end{enumerate}

We note that 2 above is characteristic of Gibbs samplers applied to
large datasets~\citep{dupuy2016}. Such samplers represent a
model posterior via counts over an entire dataset of $N$ samples. When
a minibatch of $M$ samples is used to update the model, the counts for
these samples only are updated. This creates ``steps'' of $O(M/N)$
in the model parameters, and correlated samples from the model
posterior. Correlated samples are still very useful in high-dimensional
ML problems with multi-model posteriors since they correspond to a finer-scale
random walk through the posterior landscape. 
\begin{comment}
It is worth discussing at this point the typical goals of M-H sampling on large
datasets. By the Bernstein-von Mises Theorem, the posterior distribution for a
Bayesian inference task has variance that scales inversely with $N$. Simply
sampling from it is one application, but an efficient
proposal~\citep{OptimalScaling01} has similar variance to the target and will
diffuse to it extremely slowly. For applications to neural networks or models
where the posterior is multimodal~\citep{choromanska2014loss}, samplers will
likely get trapped in one of the modes. A common solution is to anneal the
sampler, running first at high temperatures to flatten the likelihood landscape.
This in turn reduces the variance of the log acceptance probability and allows
our test to be applied. These samples can cover the search space densely with
small steps rather than taking a few sparse steps towards an optimum. In this
mode, M-H tests can be used in similar fashion to Stochastic Gradient Descent.
The goal in SGD is to make gradual progress to a posterior mode with each step,
taking small steps so that the cumulative displacement has progressively lower
variance.
\end{comment}
The contributions of this paper are as follows:

@@

    \item We compare our new test and prior approaches on several
      datasets. We demonstrate several order-of-magnitude improvements in sample efficiency,
      and that batch size distribution is short-tailed. 
\end{itemize}

@@

A priority is to extend our work to methods such as Hamiltonian Monte Carlo and Langevin Dynamics
which use efficient but asymmetric proposals. While there are various approaches
to symmetrizing these proposals, they have high cost in the context of minibatch
MCMC. Instead we plan to extend our method to log proposal ratios which have
similar structure (whole-dataset mean plus additive noise) to the log probability
ratio. These can be similarly absorbed in the Barker test.
